---
title: "Getting Started with dplR"
author: "Andy Bunn"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with dplR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dev = "png", dpi = 96
)
```

This vignette walks through the core of a tree-ring analysis in
dplR: reading ring widths, describing them, detrending, building a
chronology, and checking the crossdating. It uses only dplR and base
R, and every example runs on data that ship with the package. For a
longer treatment, with more on each step and on signal-free
chronologies and time-series analysis, see
[Learning to Love dplR](https://opendendro.github.io/dplR-workshop/).

```{r load}
library(dplR)
```

## Reading ring widths

Ring widths are read with `read.rwl()`, which handles Tucson
(decadal), compact, Heidelberg, CSV-style spreadsheets and TRiDaS
files. It guesses the format, but naming it is safer:

```{r read, eval=FALSE}
dat <- read.rwl("mysite.rwl", format = "tucson")
```

The result is an `rwl` object: a data frame with one column per
series and one row per year, with the years as row names and `NA`
where a series has no ring. Here we use `co021`, 35 Douglas-fir
series from Mesa Verde, Colorado, which was read from the
International Tree-Ring Data Bank this way.

```{r data}
data(co021)
class(co021)
dim(co021)
co021[1:5, 1:4]
```

## Describing the data

`rwl.report()` gives an overview: number of series, span, mean
length, mean interseries correlation, and any missing rings or
suspicious values.

```{r report}
rwl.report(co021)
```

`summary()` returns per-series statistics, and `plot()` shows
where each series sits in time.

```{r describe}
head(summary(co021))
plot(co021, plot.type = "spag")
```

## Detrending

Raw ring widths carry an age-related growth trend and differences in
mean growth between trees. Detrending fits a curve to each series and
divides the ring widths by it, giving dimensionless indices with a
mean of about one. `detrend()` does this for every series;
`detrend.series()` does one and plots the fit, which is a good way to
choose a method.

```{r detrend-one}
x <- co021[, "641114"]
names(x) <- rownames(co021)
x.rwi <- detrend.series(x, method = c("Spline", "ModNegExp"),
                        make.plot = TRUE)
```

Here we use a cubic smoothing spline with a 50% frequency cutoff at
two-thirds of each series' length (the default for `"Spline"`).

```{r detrend}
co021.rwi <- detrend(co021, method = "Spline")
```

The result is an `rwi` object with the same shape as the `rwl`, which
records how it was made. `summary()` describes the indices as a
collection. `rbar.eff` is the mean interseries correlation and EPS the
expressed population signal; an EPS above about 0.85 is the usual rule
of thumb for a chronology that represents the population. These come
from `rwi.stats()`, and here count each core as its own tree; pass
`ids` to group cores by tree. The summary also correlates each series
with the mean of the others and lists any that do not fit.

```{r rwi-summary}
summary(co021.rwi)
```

`plot()` with `plot.type = "image"` shows every index at once, years
across and series down, brown below 1 and green above. Vertical stripes
are years the trees agree on, which is the signal a chronology is built
from.

```{r rwi-image, fig.height=6}
plot(co021.rwi, plot.type = "image")
```

The same plot is a quick check on the detrending. Dividing each series
by its mean leaves the age trend in, and it shows as a green run at the
start of nearly every series.

```{r rwi-image-mean, fig.height=6}
plot(detrend(co021, method = "Mean"), plot.type = "image")
```

## Building a chronology

`chron()` averages the indices by year, using Tukey's biweight robust
mean by default. The result is a `crn` object: the chronology and the
number of series behind each year.

```{r chron}
co021.crn <- chron(co021.rwi)
tail(co021.crn)
plot(co021.crn, add.spline = TRUE, nyrs = 32)
```

The early part of this chronology rests on few series (fewer than
five before 1234), so check `samp.depth` before trusting a given year.

## Checking the crossdating

Crossdating assigns each ring its exact calendar year. dplR does not
replace visual crossdating, but it can check it statistically, the way
COFECHA does. To see what an error looks like, we plant one: the
1500 ring of series 641143 is deleted, so every ring before 1500 is
now dated one year too late.

```{r plant}
dat <- co021
x <- dat[, "641143"]
names(x) <- rownames(dat)
dat[, "641143"] <- delete.ring(x, year = 1500)
```

`corr.rwl.seg()` correlates overlapping segments of each series
against a master built from all the other series. Segments that do not
correlate significantly are flagged, and with `lag.max` it also finds
the shift at which each segment correlates best.

```{r crs, fig.height=6}
crs <- corr.rwl.seg(dat, seg.length = 50, pcrit = 0.01,
                    lag.max = 10, label.cex = 0.7)
```

Every tested segment of 641143 that ends before 1500 correlates best
at a lag of -1, and every later one at a lag of 0.
In dplR, as in COFECHA, a negative lag means the series is missing a
ring, so this points to a missing ring near 1500.

```{r best-lag}
ok <- !is.na(crs$best.lag["641143", ])
crs$best.lag["641143", ok]
```

`ccf.series.rwl()` looks at one series in more detail, plotting the
full cross-correlation with the master for each segment.

```{r ccf, fig.height=6}
ccf <- ccf.series.rwl(rwl = dat[, colnames(dat) != "641143"],
                      series = dat[, "641143"],
                      series.yrs = as.numeric(rownames(dat)),
                      seg.length = 50, bin.floor = 50)
```

`xdate.report()` puts all of this into a COFECHA-style report. It
lists the flagged segments with their best lag and the gain in
correlation, and records the settings and file checksum so the report
can be reproduced. Printing it shows the report in the console.

```{r xdate-report, results="hide"}
rpt <- xdate.report(dat, title = "co021 with a planted fault")
rpt
```

```{r report-sections, include=FALSE}
## Split the Markdown version of the report into its sections so the
## vignette can show a couple of them as tables. Headings become bold
## labels, which keeps them out of the table of contents and out of
## pandoc's section structure.
md <- format(rpt, type = "markdown")
md <- sub("^#{1,2} (.*)$", "**\\1**", md)
sec <- cumsum(grepl("^\\*\\*", md))
sections <- split(md, sec)
names(sections) <- sapply(sections, function(s) gsub("\\*", "", s[1]))
show.section <- function(name) {
    cat(sections[[name]], sep = "\n")
    cat("\n")
}
```

The report opens with a summary of the collection:

```{r report-summary, echo=FALSE, results="asis"}
show.section("Summary")
```

The flagged segments section is where to start. All nine flags are on
641143, all are `B` flags at a lag of -1, and all end before 1500:

```{r report-flags, echo=FALSE, results="asis"}
show.section("Flagged segments")
```

The same table is in the report as a data frame, `rpt$flagged`, ready
for further work:

```{r flagged}
rpt$flagged[, c("series", "from", "to", "flag", "best.lag", "gain")]
```

A `B` flag means the segment correlates better at another lag than at
its dated position. An `A` flag, which does not appear here, means the
dated position is the segment's best but falls short of significance.
The report also has a correlation table for every segment of every
series, descriptive statistics, the output of `rwl.check()`, and notes
on how each number was computed:

<details>
<summary>The full report</summary>

<div style="overflow-x: auto;">
```{r report-full, echo=FALSE, results="asis"}
cat(md, sep = "\n")
```
</div>

</details>

`write.xdate.report()` saves the report as text, Markdown or HTML:

```{r write-report, eval=FALSE}
write.xdate.report(rpt, "co021-report.html")
```

The statistics say where to look; the wood says what happened. Having
found the likely missing ring, you would go back to the sample, and
once confirmed, fix the series with `insert.ring()`.

## Where to go next

- [Learning to Love dplR](https://opendendro.github.io/dplR-workshop/)
  covers these steps and more in depth.
- `vignette("math-dplR")` gives the mathematics behind the smoothing
  splines, detrending and other functions.
- `help(package = "dplR")` lists every function. Good next stops are
  `?detrend`, `?chron`, `?ssf` (signal-free chronologies),
  `?corr.rwl.seg` and `?xdate.report`.
- dplR is part of [openDendro](https://opendendro.org).
